---
title: "Telecom Customer Churn Prediction Project Proposal"
author: "Ashish Yakub Beary"
date: "`r Sys.Date()`"
output: 
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load necessary packages
library(tidyverse)
library(janitor)
library(skimr)
library(knitr)
library(caret)
library(corrplot)
library(randomForest)
library(pROC)
```

## Research Question

* What factors most significantly predict customer churn in the telecom industry?
* Secondary questions:
  * How do service usage patterns and demographics correlate with churn probability?
  * Which specific services or contract features have the strongest protective effect against churn?
  * Can we identify high-risk customers early to implement targeted retention strategies?

## Cases

* Each case represents an individual customer of a telecommunications company
* The dataset contains 7,043 unique customers
* Each record includes comprehensive information about customer demographics, service subscriptions, billing information, and churn status

## Method of Data Collection

* The data was collected by the telecommunications company from their customer relationship management (CRM) system and billing databases
* It represents historical customer data from account creation through either current active status or account termination
* The information includes service subscriptions, billing details, demographic information, and geographic data

## Type of Study

* This is an observational study as it analyzes existing customer data without experimental manipulation
* The study examines patterns and relationships in historical data to identify predictive factors for customer churn

## Data Source

* Primary dataset: telecom_customer_churn.csv (7,043 rows × 38 columns)
* Supplementary datasets: 
  * telecom_zipcode_population.csv - population data by zip code
  * telecom_data_dictionary.csv - metadata describing each variable

## Variables

### Response Variable

* **Customer Status** (categorical): Customer's current status, with "Churned" indicating customers who have left the company
* For binary analysis purposes, this will be transformed into a churn indicator (Yes/No)

### Explanatory Variables

#### Demographics
* Gender (categorical)
* Age (numerical)
* Married (categorical - Yes/No)
* Number of Dependents (numerical)
* City (categorical)
* Zip Code (categorical)
* Geographic location (Latitude/Longitude)

#### Account Information
* Number of Referrals (numerical)
* Tenure in Months (numerical)
* Offer (categorical)
* Contract (categorical - Month-to-Month, One Year, Two Year)
* Paperless Billing (categorical - Yes/No)
* Payment Method (categorical)

#### Service Subscriptions
* Phone Service (categorical - Yes/No)
* Multiple Lines (categorical)
* Internet Service (categorical - Yes/No)
* Internet Type (categorical)
* Avg Monthly GB Download (numerical)
* Online Security (categorical - Yes/No)
* Online Backup (categorical - Yes/No)
* Device Protection Plan (categorical - Yes/No)
* Premium Tech Support (categorical - Yes/No)
* Streaming TV (categorical - Yes/No)
* Streaming Movies (categorical - Yes/No)
* Streaming Music (categorical - Yes/No)
* Unlimited Data (categorical - Yes/No)

#### Financial Metrics
* Monthly Charge (numerical)
* Total Charges (numerical)
* Total Refunds (numerical)
* Total Extra Data Charges (numerical)
* Total Long Distance Charges (numerical)
* Total Revenue (numerical)
* Avg Monthly Long Distance Charges (numerical)

#### Churn Details (for churned customers only)
* Churn Category (categorical)
* Churn Reason (categorical)

## Preliminary Data Analysis

```{r load-data, eval=TRUE}
# Load the datasets
telecom_churn <- read.csv("telecom_customer_churn.csv", stringsAsFactors = TRUE)
zipcode_population <- read.csv("telecom_zipcode_population.csv")
data_dictionary <- read.csv("telecom_data_dictionary.csv", encoding = "CP1252")

# Clean column names
telecom_churn <- clean_names(telecom_churn)
zipcode_population <- clean_names(zipcode_population)
data_dictionary <- clean_names(data_dictionary)

# Display dataset structure 
str(telecom_churn)

# Create binary churn variable for analysis
telecom_churn$churned <- ifelse(telecom_churn$customer_status == "Churned", "Yes", "No")
telecom_churn$churned <- as.factor(telecom_churn$churned)

# Basic summary statistics
summary(telecom_churn[c("age", "tenure_in_months", "number_of_dependents", 
                         "avg_monthly_gb_download", "monthly_charge", "total_charges")])
```

```{r eda-churn-visualization, fig.width=10, fig.height=6}
# Calculate overall churn rate
churn_rate <- mean(telecom_churn$customer_status == "Churned") * 100

# Visualize the churn distribution
ggplot(telecom_churn, aes(x = customer_status)) +
  geom_bar(fill = c("darkgreen", "darkred", "steelblue")) +
  geom_text(stat = "count", aes(label = scales::percent(..count../sum(after_stat(count)))), 
            vjust = -0.5) +
  labs(title = paste0("Customer Status Distribution (Churn Rate: ", round(churn_rate, 1), "%)"),
       x = "Customer Status",
       y = "Count") +
  theme_minimal()
```

```{r categorical-analysis, fig.width=12, fig.height=8}
# Analyze categorical variables by churn status
cat_vars <- c("contract", "internet_type", "payment_method", "offer", "paperless_billing")

# Create a function to plot churn rate by category
plot_churn_by_category <- function(data, variable) {
  # Calculate percentages
  churn_by_cat <- data %>%
    group_by(!!sym(variable)) %>%
    summarize(
      total = n(),
      churned = sum(customer_status == "Churned"),
      churn_rate = churned / total * 100
    ) %>%
    arrange(desc(churn_rate))
  
  # Create plot
  ggplot(churn_by_cat, aes(x = reorder(!!sym(variable), -churn_rate), y = churn_rate)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    geom_text(aes(label = paste0(round(churn_rate, 1), "%")), vjust = -0.5) +
    labs(title = paste("Churn Rate by", variable),
         x = variable,
         y = "Churn Rate (%)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Plot for contract type
plot_churn_by_category(telecom_churn, "contract")
```

```{r numerical-analysis, fig.width=10, fig.height=6}
# Analyze numerical variables by churn status
num_vars <- c("age", "tenure_in_months", "monthly_charge", "total_charges", "avg_monthly_gb_download")

# Create summary statistics by churn status
num_summary <- telecom_churn %>%
  group_by(churned) %>%
  summarize(
    avg_age = mean(age, na.rm = TRUE),
    avg_tenure = mean(tenure_in_months, na.rm = TRUE),
    avg_monthly_charge = mean(monthly_charge, na.rm = TRUE),
    avg_total_charges = mean(total_charges, na.rm = TRUE),
    avg_monthly_download = mean(avg_monthly_gb_download, na.rm = TRUE)
  )

print(num_summary)

# Visualize tenure by churn status
ggplot(telecom_churn, aes(x = churned, y = tenure_in_months, fill = churned)) +
  geom_boxplot() +
  labs(title = "Customer Tenure by Churn Status",
       x = "Churned",
       y = "Tenure (months)") +
  scale_fill_manual(values = c("darkgreen", "darkred")) +
  theme_minimal()
```

```{r service-adoption, fig.width=12, fig.height=8}
# Analyze service adoption and its impact on churn
service_vars <- c("phone_service", "multiple_lines", "online_security", 
                  "online_backup", "device_protection_plan", "premium_tech_support",
                  "streaming_tv", "streaming_movies", "streaming_music", "unlimited_data")

# Specify which services to include in the visualization
services_to_plot <- c("online_security", "premium_tech_support", "contract")

# Create a function to calculate and visualize service impact on churn
service_impact <- function(data, service_var) {
  # Filter out NA values
  data_filtered <- data %>% filter(!is.na(!!sym(service_var)))
  
  # Calculate churn rates
  service_churn <- data_filtered %>%
    group_by(!!sym(service_var)) %>%
    summarize(
      total = n(),
      churned = sum(customer_status == "Churned"),
      churn_rate = churned / total * 100
    )
  
  return(service_churn)
}

# Example for one service
online_security_impact <- service_impact(telecom_churn, "online_security")
print(online_security_impact)

# Create an empty data frame with the right structure
services_plot_data <- data.frame(
  service = character(),
  service_value = character(),
  total = numeric(),
  churned = numeric(),
  churn_rate = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each service
for (service_name in services_to_plot) {
  # Get the churn data for this service
  service_data <- service_impact(telecom_churn, service_name)
  
  # Extract the service value column (which has a dynamic name)
  service_values <- service_data[[1]]  # The first column contains the service values
  
  # Create a new data frame with consistent column names
  temp_df <- data.frame(
    service = service_name,
    service_value = as.character(service_values),
    total = service_data$total,
    churned = service_data$churned,
    churn_rate = service_data$churn_rate,
    stringsAsFactors = FALSE
  )
  
  # Add to the main data frame
  services_plot_data <- rbind(services_plot_data, temp_df)
}

# Plot with the corrected data structure
ggplot(services_plot_data, aes(x = reorder(paste(service, service_value), -churn_rate), y = churn_rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = paste0(round(churn_rate, 1), "%")), vjust = -0.5) +
  labs(title = "Churn Rate by Selected Services",
       x = "Service and Status",
       y = "Churn Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Relevant Summary Statistics

Based on the preliminary analysis, the following summary statistics are relevant for understanding customer churn:

* Overall churn rate: Approximately 26.5% of customers have churned
* Demographic statistics:
  * Age distribution shows typical consumer age range (18-80 years)
  * Geographic distribution across multiple cities and zip codes
* Service adoption statistics:
  * Internet service types (Fiber Optic, DSL, None)
  * Additional service adoption rates (security, backup, streaming, etc.)
* Financial metrics:
  * Average monthly charges for churned vs. retained customers
  * Total charges and revenue differences between customer groups
* Contract and tenure statistics:
  * Contract type distribution shows higher churn for month-to-month contracts
  * Average tenure for churned customers is significantly lower (approximately 18 months vs. 38 months for non-churned)

## Statistical Methods

### Primary Analysis Method: Logistic Regression

Logistic regression is appropriate for this analysis because:

* The response variable (churn) is binary (Yes/No)
* We need to quantify the effect of multiple predictors on churn probability
* We want to obtain interpretable odds ratios for business decision-making
* It can handle both categorical and numerical predictors


```{r logistic-regression-example, eval=TRUE}
# Example of logistic regression model (simplified)
churn_model <- glm(
  churned ~ contract + internet_type + tenure_in_months + monthly_charge + 
           online_security + premium_tech_support,
  family = binomial(link = "logit"),
  data = telecom_churn
)

# Model summary
summary(churn_model)

# Example prediction
predicted_probs <- predict(churn_model, type = "response")
telecom_churn$predicted_churn_prob <- predicted_probs

# ROC curve assessment
roc_obj <- roc(telecom_churn$churned, predicted_probs)
auc_value <- auc(roc_obj)
cat("AUC:", auc_value)
```

### Secondary Analysis Methods

1. **Random Forest Classification**
   * Will help identify complex non-linear relationships and interactions
   * Provides feature importance to highlight the most predictive variables
   * Handles mixed data types effectively

```{r random-forest-example, eval=FALSE}
# Random Forest example (not evaluated to save computation time)
set.seed(123)
rf_model <- randomForest(
  churned ~ contract + internet_type + tenure_in_months + monthly_charge + 
            online_security + premium_tech_support + payment_method + age,
  data = telecom_churn,
  ntree = 100,
  importance = TRUE
)

# Variable importance
varImpPlot(rf_model)
```

2. **Survival Analysis**
   * Can analyze time-to-churn based on tenure
   * Provides insights into when customers are most at risk of churning
   * Allows for censored observations (current customers who haven't churned yet)

### Model Evaluation Strategy

The models will be evaluated using:

* Train/test split (70%/30%) for model validation
* Cross-validation to ensure model robustness
* ROC curves and AUC for classification performance
* Confusion matrix for precision, recall, and F1-score
* McFadden's R² for logistic regression fit assessment

## Expected Outcomes

This analysis is expected to:

1. Identify the key predictors of customer churn in the telecom industry
2. Quantify the impact of each factor on churn probability
3. Develop a predictive model to identify at-risk customers before they churn
4. Provide actionable insights for reducing churn through targeted interventions
5. Generate recommendations for service improvements and retention strategies

The results will be valuable for:

* Marketing teams designing retention campaigns
* Product managers prioritizing service improvements
* Customer service teams implementing proactive retention measures
* Business leaders making strategic decisions about service offerings