---
title: "Telecom Customer Churn Prediction Analysis"
author: "Ashish Yakub Beary"
date: "2025-04-20"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: kate
    code_folding: show
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 10
)
```

# 1. Introduction

This report presents an analysis of customer churn for a telecommunications company. Customer churn, or the rate at which customers stop doing business with a company, is a critical metric in the telecom industry due to the high cost of acquiring new customers compared to retaining existing ones.

## 1.1 Research Questions

The primary research question addressed in this analysis is:

* What factors most significantly predict customer churn in the telecom industry?

Secondary questions include:

* How do service usage patterns and demographics correlate with churn probability?
* Which specific services or contract features have the strongest protective effect against churn?
* Can we identify high-risk customers early to implement targeted retention strategies?

## 1.2 Dataset Overview

The analysis is based on the following datasets:

* **telecom_customer_churn.csv**: Primary dataset containing customer information (7,043 customers with 38 variables)
* **telecom_zipcode_population.csv**: Supplementary dataset with population information by zip code
* **telecom_data_dictionary.csv**: Metadata describing each variable

```{r load-libraries}
# Load necessary packages
library(tidyverse)  # For data manipulation and visualization
library(caret)      # For machine learning workflow
library(randomForest) # For random forest model
library(pROC)       # For ROC curve analysis
library(corrplot)   # For correlation visualization
library(janitor)    # For cleaning column names
library(scales)     # For nice scales on plots
library(knitr)      # For tables
library(kableExtra) # For enhanced tables
library(viridis)    # For nice color palettes
library(gridExtra)  # For combining plots
library(pdp)        # For partial dependence plots
library(broom)

# Set seed for reproducibility
set.seed(123)
```

# 2. Data Loading and Initial Exploration

```{r load-data}
# Read the datasets
telecom_churn <- read.csv("telecom_customer_churn.csv", stringsAsFactors = TRUE)
zipcode_population <- read.csv("telecom_zipcode_population.csv")
data_dictionary <- read.csv("telecom_data_dictionary.csv", encoding = "CP1252")

# Clean column names
telecom_churn <- clean_names(telecom_churn)
zipcode_population <- clean_names(zipcode_population)
data_dictionary <- clean_names(data_dictionary)
```

## 2.1 Data Structure

Let's examine the structure of our main dataset:

```{r data-structure}
# Display the structure of the first few columns
str(telecom_churn[,])

# Get a summary of the dataset dimensions
cat("Number of customers:", nrow(telecom_churn), "\n")
cat("Number of variables:", ncol(telecom_churn), "\n")
```

## 2.2 Create Binary Churn Variable

For our analysis, we'll create a binary churn variable that indicates whether a customer has churned or not.

```{r binary-churn}
# Create binary churn variable for analysis
telecom_churn$churned <- ifelse(telecom_churn$customer_status == "Churned", "Yes", "No")
telecom_churn$churned <- as.factor(telecom_churn$churned)

# Check distribution
table(telecom_churn$churned)
```

## 2.3 Basic Summary Statistics

```{r summary-stats}
# Basic summary statistics for key numerical variables
summary(telecom_churn[c("age", "tenure_in_months", "number_of_dependents",
                       "avg_monthly_gb_download", "monthly_charge", "total_charges")])
```

# 3. Data Cleaning and Preprocessing

## 3.1 Missing Values

```{r missing-values}
# Check for missing values
missing_values <- colSums(is.na(telecom_churn))
missing_values[missing_values > 0]

# Handle missing values for avg_monthly_gb_download using median imputation
telecom_churn$avg_monthly_gb_download[is.na(telecom_churn$avg_monthly_gb_download)] <- 
  median(telecom_churn$avg_monthly_gb_download, na.rm = TRUE)

telecom_churn$avg_monthly_long_distance_charges[is.na(telecom_churn$avg_monthly_long_distance_charges)] <- 
  median(telecom_churn$avg_monthly_long_distance_charges, na.rm = TRUE)
```

## 3.2 Handling Categorical Variables

Some categorical variables have empty values because they are conditionally relevant. For example, internet-related services are only applicable to customers with internet service. 

```{r handle-categorical}
# Convert empty strings to NA for certain categorical columns
service_cols <- c("multiple_lines", "internet_type", "online_security", 
                 "online_backup", "device_protection_plan", "premium_tech_support",
                 "streaming_tv", "streaming_movies", "streaming_music", "unlimited_data")

for(col in service_cols) {
  telecom_churn[[col]] <- as.character(telecom_churn[[col]])
  telecom_churn[[col]][telecom_churn[[col]] == ""] <- NA
  telecom_churn[[col]] <- as.factor(telecom_churn[[col]])
}

# Some customers don't have internet service, which is why they have NA for internet-related services
# We'll recode these NAs as "No Internet Service"
internet_related <- c("internet_type", "online_security", "online_backup", 
                      "device_protection_plan", "premium_tech_support",
                      "streaming_tv", "streaming_movies", "streaming_music", 
                      "unlimited_data")

for(col in internet_related) {
  levels(telecom_churn[[col]]) <- c(levels(telecom_churn[[col]]), "No Internet Service")
  telecom_churn[[col]][is.na(telecom_churn[[col]]) & telecom_churn$internet_service == "No"] <- "No Internet Service"
}

# Similarly for phone-related services
phone_related <- c("multiple_lines")
for(col in phone_related) {
  levels(telecom_churn[[col]]) <- c(levels(telecom_churn[[col]]), "No Phone Service")
  telecom_churn[[col]][is.na(telecom_churn[[col]]) & telecom_churn$phone_service == "No"] <- "No Phone Service"
}

```

# 4. Exploratory Data Analysis

## 4.1 Overall Churn Rate

```{r churn-rate}
# Calculate overall churn rate
churn_rate <- mean(telecom_churn$customer_status == "Churned") * 100
cat("Overall churn rate:", round(churn_rate, 2), "%\n")

# Visualize the churn distribution
ggplot(telecom_churn, aes(x = customer_status)) +
  geom_bar(fill = c("#52664b", "#2e6083", "#2e6083")) +
  geom_text(stat = "count", aes(label = scales::percent(after_stat(count)/sum(after_stat(count)))),
            vjust = -0.5) +
  labs(title = paste0("Customer Status Distribution (Churn Rate: ", round(churn_rate, 1), "%)"),
       x = "Customer Status",
       y = "Count") +
  theme_minimal()
```

## 4.2 Numeric Variables and Churn

Let's examine the relationship between key numerical variables and churn:

```{r numeric-vars}
# Age distribution by churn status
p1 <- ggplot(telecom_churn, aes(x = churned, y = age, fill = churned)) +
  geom_boxplot() +
  labs(title = "Age Distribution by Churn Status",
       x = "Churned",
       y = "Age") +
  scale_fill_manual(values = c("#2e6083", "#52664b")) +
  theme_minimal()

# Tenure distribution by churn status
p2 <- ggplot(telecom_churn, aes(x = churned, y = tenure_in_months, fill = churned)) +
  geom_boxplot() +
  labs(title = "Customer Tenure by Churn Status",
       x = "Churned",
       y = "Tenure (months)") +
  scale_fill_manual(values = c("#2e6083", "#52664b")) +
  theme_minimal()

# Monthly charge distribution by churn status
p3 <- ggplot(telecom_churn, aes(x = churned, y = monthly_charge, fill = churned)) +
  geom_boxplot() +
  labs(title = "Monthly Charge by Churn Status",
       x = "Churned",
       y = "Monthly Charge ($)") +
  scale_fill_manual(values = c("#2e6083", "#52664b")) +
  theme_minimal()

# Display plots side by side
grid.arrange(p1, p2, p3, ncol = 3)
```

```{r numeric-summary}
# Calculate mean statistics by churn status
telecom_churn %>%
  group_by(churned) %>%
  summarize(
    avg_age = mean(age, na.rm = TRUE),
    avg_tenure = mean(tenure_in_months, na.rm = TRUE),
    avg_monthly_charge = mean(monthly_charge, na.rm = TRUE),
    avg_total_charges = mean(total_charges, na.rm = TRUE),
    avg_monthly_download = mean(avg_monthly_gb_download, na.rm = TRUE)
  ) %>%
  kable(caption = "Key Metrics by Churn Status", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Observations:**

1. **Age**: Churned customers tend to be slightly older on average.
2. **Tenure**: There's a substantial difference in tenure between churned and retained customers. Customers who churn have much shorter tenure on average.
3. **Monthly Charge**: Churned customers have higher monthly charges on average.

## 4.3 Categorical Variables and Churn

Let's analyze how categorical variables relate to churn:

```{r categorical-vars}
# Create a function to plot churn rate by category
plot_churn_by_category <- function(data, variable) {
  # Calculate percentages
  churn_by_cat <- data %>%
    group_by(!!sym(variable)) %>%
    summarize(
      total = n(),
      churned = sum(customer_status == "Churned"),
      churn_rate = churned / total * 100
    ) %>%
    arrange(desc(churn_rate))
  
  # Create plot
  ggplot(churn_by_cat, aes(x = reorder(!!sym(variable), -churn_rate), y = churn_rate)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    geom_text(aes(label = paste0(round(churn_rate, 1), "%")), vjust = -0.5) +
    labs(title = paste("Churn Rate by", variable),
         x = variable,
         y = "Churn Rate (%)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Plot for contract type
plot_churn_by_category(telecom_churn, "contract")
```

```{r internet-type}
# Plot for internet type
plot_churn_by_category(telecom_churn, "internet_type")
```

```{r payment-method}
# Plot for payment method
plot_churn_by_category(telecom_churn, "payment_method")
```

**Observations:**

1. **Contract Type**: Month-to-month contracts have a significantly higher churn rate (45.8%) compared to one-year (10.7%) and two-year contracts (2.5%).
2. **Internet Type**: Fiber optic internet customers have the highest churn rate, while DSL customers and customers opting out of internet service have a lower churn rate.
3. **Payment Method**: Customers using electronic checks as their payment method have a higher churn rate compared to other payment methods.

## 4.4 Service Adoption Impact on Churn

Let's analyze how various services impact churn:

```{r service-impact}
# Create a function to calculate service impact on churn
service_impact <- function(data, service_var) {
  # Calculate churn rates
  service_churn <- data %>%
    group_by(!!sym(service_var)) %>%
    summarize(
      total = n(),
      churned = sum(customer_status == "Churned"),
      churn_rate = churned / total * 100
    )
  
  return(service_churn)
}

# Example for online security
online_security_impact <- service_impact(telecom_churn, "online_security")
kable(online_security_impact, caption = "Churn Rate by Online Security Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Let's analyze multiple services together:

```{r multiple-services}
# Selected services to analyze
selected_services <- c("online_security", "premium_tech_support", "contract")

# Create an empty data frame with the right structure
services_plot_data <- data.frame(
  service = character(),
  service_value = character(),
  total = numeric(),
  churned = numeric(),
  churn_rate = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each service
for (service_name in selected_services) {
  # Get the churn data for this service
  service_data <- service_impact(telecom_churn, service_name)
  
  # Extract the service value column (which has a dynamic name)
  service_values <- service_data[[1]] # The first column contains the service values
  
  # Create a new data frame with consistent column names
  temp_df <- data.frame(
    service = service_name,
    service_value = as.character(service_values),
    total = service_data$total,
    churned = service_data$churned,
    churn_rate = service_data$churn_rate,
    stringsAsFactors = FALSE
  )
  
  # Add to the main data frame
  services_plot_data <- rbind(services_plot_data, temp_df)
}

# Plot with the data
ggplot(services_plot_data, aes(x = reorder(paste(service, service_value), -churn_rate), y = churn_rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = paste0(round(churn_rate, 1), "%")), vjust = -0.5) +
  labs(title = "Churn Rate by Selected Services",
       x = "Service and Status",
       y = "Churn Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Observations:**

1. Customers without online security have a much higher churn rate (41.8%) than those with online security (14.6%).
2. Similar patterns are observed for premium tech support.
3. This suggests that additional services may create "stickiness" and reduce churn.

## 4.5 Correlation Analysis

Let's examine the correlations between numerical variables:

```{r correlation}
# Select numerical columns for correlation analysis
numeric_cols <- telecom_churn %>%
  select(age, tenure_in_months, number_of_dependents, avg_monthly_long_distance_charges,
         avg_monthly_gb_download, monthly_charge, total_charges, total_refunds,
         total_extra_data_charges, total_long_distance_charges, total_revenue) %>%
  names()

# Calculate correlation matrix
correlation_matrix <- cor(telecom_churn[numeric_cols], use = "pairwise.complete.obs")

# Create a correlation plot
corrplot(correlation_matrix, method = "circle", type = "upper", 
         tl.col = "black", tl.srt = 45, tl.cex = 0.7,
         title = "Correlation Matrix of Numerical Variables",
         mar = c(0, 0, 1, 0))
```

**Key Correlations:**

1. Tenure is positively correlated with total charges, as expected.
2. Monthly charge is positively correlated with average monthly GB download.
3. Total revenue is strongly correlated with total charges, monthly charge, and tenure. Moderately with long distance charges.

# 5. Feature Engineering

## 5.1 Create Derived Features

Let's create some additional features that might help improve our predictive models:

```{r feature-engineering}
# Customer lifetime value (CLV)
telecom_churn$customer_lifetime_value <- telecom_churn$total_revenue / telecom_churn$tenure_in_months

# Average monthly revenue
telecom_churn$avg_monthly_revenue <- telecom_churn$total_revenue / telecom_churn$tenure_in_months

# Service count (number of additional services subscribed)
telecom_churn$service_count <- rowSums(telecom_churn[, c("online_security", 
                                                       "online_backup", 
                                                       "device_protection_plan", 
                                                       "premium_tech_support",
                                                       "streaming_tv", 
                                                       "streaming_movies", 
                                                       "streaming_music")] == "Yes", na.rm = TRUE)

# Visualize service count distribution by churn status
ggplot(telecom_churn, aes(x = service_count, fill = churned)) +
  geom_histogram(position = "dodge", binwidth = 1) +
  labs(title = "Service Count Distribution by Churn Status",
       x = "Number of Services",
       y = "Count") +
  scale_fill_manual(values = c( "#2e6083", "#52664b")) +
  theme_minimal()
```

## 5.2 Add Population Data

Let's join the zipcode population data to potentially identify demographic patterns:

```{r join-population}
# Join zipcode population data
telecom_churn <- left_join(telecom_churn, zipcode_population, by = "zip_code")

# Calculate population density quartiles
telecom_churn$population_quartile <- ntile(telecom_churn$population, 4)

# Analyze churn by population quartile
telecom_churn %>%
  group_by(population_quartile) %>%
  summarize(
    churn_rate = mean(churned == "Yes") * 100,
    customer_count = n()
  ) %>%
  kable(caption = "Churn Rate by Population Quartile", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# 6. Modeling Preparation

## 6.1 Encoding Categorical Variables

For our models, we need to encode categorical variables:

```{r encoding}
# For variables with more than two levels, create dummy variables
dummy_vars <- c("contract", "internet_type", "payment_method", "offer")

# Create a formula for model matrix
dummy_formula <- as.formula(paste("~", paste(dummy_vars, collapse = " + ")))

# Generate dummy variables
dummy_data <- model.matrix(dummy_formula, data = telecom_churn)[, -1] # Remove intercept

# Check the actual column names in the dummy data
colnames(dummy_data)[1:10]  # Look at the first 10 column names to understand the naming pattern

# Combine with original data
telecom_churn_encoded <- cbind(telecom_churn, as.data.frame(dummy_data))

# Use the correct column names when checking the data
# For example, if the proper names are:
head(telecom_churn_encoded[, grep("contract", colnames(telecom_churn_encoded), value = TRUE)])
```

## 6.2 Feature Selection

Let's select relevant features for our models:

```{r feature-selection}
# Exclude redundant or irrelevant columns
exclude_cols <- c("customer_id", "customer_status", "churn_category", "churn_reason", 
                 "latitude", "longitude", "zip_code", "city")

# Create a new data frame with selected features
model_data <- telecom_churn_encoded %>%
  select(-one_of(exclude_cols))

# Check class balance
table(model_data$churned)
```

## 6.3 Data Splitting

Let's split our data into training and testing sets:

```{r data-splitting}
# Split the data into training and testing sets (70/30)
train_index <- createDataPartition(model_data$churned, p = 0.7, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

# Check dimensions
cat("Training set dimensions:", dim(train_data), "\n")
cat("Testing set dimensions:", dim(test_data), "\n")

# Check class balance in training set
table(train_data$churned)
```

# 7. Logistic Regression Model

## 7.1 Model Building

Let's build a logistic regression model using key variables from our EDA:

```{r logistic-regression}
# Select key variables for the first model based on our EDA
logistic_vars <- c("tenure_in_months", "contract", "monthly_charge", 
                  "internet_type", "online_security", "premium_tech_support",
                  "payment_method", "paperless_billing", "service_count",
                  "customer_lifetime_value")

# Create formula for logistic regression
logistic_formula <- as.formula(paste("churned ~", paste(logistic_vars, collapse = " + ")))

# Fit logistic regression model
logistic_model <- glm(logistic_formula, family = binomial(link = "logit"), data = train_data)

# Model summary
summary(logistic_model)
```

## 7.2 Odds Ratios

Let's examine the odds ratios to interpret the model coefficients:

```{r odds-ratios}
# Calculate odds ratios
odds_ratios <- exp(coef(logistic_model))
odds_ratios_df <- data.frame(
  Variable = names(odds_ratios),
  OddsRatio = odds_ratios,
  LowerCI = exp(confint(logistic_model))[, 1],
  UpperCI = exp(confint(logistic_model))[, 2]
)

# Display odds ratios
odds_ratios_df %>%
  kable(caption = "Odds Ratios for Logistic Regression Model", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

**Interpretation:**

- **Tenure**: For each additional month of tenure, the odds of churning decrease by approximately 2.6%.
- **Contract**: Compared to month-to-month contracts, one-year contracts reduce churn odds by approximately 70%, while two-year contracts reduce churn odds by approximately 90%.
- **Monthly Charge**: Higher monthly charges slightly increase churn odds (~1.1% per dollar).
- **Services**:  Customers with online security or premium tech support are significantly less likely to churn, suggesting these services enhance customer retention.
 - Online Security (Yes): OR = 0.543 → ~46% reduction in churn odds
 - Premium Tech Support (Yes): OR = 0.584 → ~42% reduction

## 7.3 Model Evaluation

Let's evaluate the logistic regression model on the test data:

```{r logistic-evaluation}
# Predict on test data
test_predictions_prob <- predict(logistic_model, newdata = test_data, type = "response")
test_predictions <- ifelse(test_predictions_prob > 0.5, "Yes", "No")

# Create confusion matrix
conf_matrix <- confusionMatrix(as.factor(test_predictions), test_data$churned, positive = "Yes")
print(conf_matrix)

# ROC curve
roc_obj <- roc(test_data$churned, test_predictions_prob)
auc_value <- auc(roc_obj)

# Plot ROC curve##

ggroc(roc_obj) +
  geom_abline(intercept = 1, slope = 1, linetype = "dashed", color = "gray") +
  labs(title = paste("ROC Curve - Logistic Regression (AUC =", round(auc_value, 3), ")"),
       x = "False Positive Rate", 
       y = "True Positive Rate") +
  theme_minimal()
```

# 8. Random Forest Model

## 8.1 Model Building

Now let's build a random forest model:

```{r random-forest}
# Prepare data for random forest (handle factors appropriately)
rf_data_train <- train_data
rf_data_test <- test_data

# Make sure the response is a factor
rf_data_train$churned <- as.factor(rf_data_train$churned)
rf_data_test$churned <- as.factor(rf_data_test$churned)

# Train random forest model
set.seed(123)
rf_model <- randomForest(
  churned ~ tenure_in_months + contract + monthly_charge + internet_type + 
    online_security + premium_tech_support + payment_method + 
    paperless_billing + service_count + avg_monthly_gb_download + 
    age + number_of_dependents,
  data = rf_data_train,
  ntree = 300,
  mtry = 5,
  importance = TRUE
)

# Model summary
print(rf_model)
```

## 8.2 Variable Importance

Let's examine which variables are most important in the random forest model:

```{r rf-importance}
# Variable importance
var_importance <- importance(rf_model)
var_importance_df <- data.frame(
  Variable = rownames(var_importance),
  MeanDecreaseGini = var_importance[, "MeanDecreaseGini"]
)
var_importance_df <- var_importance_df[order(var_importance_df$MeanDecreaseGini, decreasing = TRUE), ]

# Display top 10 variables
var_importance_df[1:10, ] %>%
  kable(caption = "Top 10 Most Important Variables in Random Forest Model", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Plot variable importance
ggplot(var_importance_df[1:15, ], aes(x = reorder(Variable, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "#52664b") +
  coord_flip() +
  labs(title = "Random Forest - Variable Importance",
       x = "Variable",
       y = "Mean Decrease in Gini") +
  theme_minimal()
```

## 8.3 Model Evaluation

Let's evaluate the random forest model on the test data:

```{r rf-evaluation}
# Predict on test data
rf_predictions <- predict(rf_model, rf_data_test, type = "class")
rf_predictions_prob <- predict(rf_model, rf_data_test, type = "prob")[, "Yes"]

# Create confusion matrix
rf_conf_matrix <- confusionMatrix(rf_predictions, rf_data_test$churned, positive = "Yes")
print(rf_conf_matrix)

# ROC curve for Random Forest
rf_roc_obj <- roc(rf_data_test$churned, rf_predictions_prob)
rf_auc_value <- auc(rf_roc_obj)

# Plot ROC curve for Random Forest
ggroc(rf_roc_obj) +
  geom_abline(intercept = 1, slope = 1, linetype = "dashed", color = "gray") +
  labs(title = paste("ROC Curve - Random Forest (AUC =", round(rf_auc_value, 3), ")"),
       x = "False Positive Rate", 
       y = "True Positive Rate") +
  theme_minimal()
```

# 9. Model Comparison

Let's compare the performance of our logistic regression and random forest models:

```{r model-comparison}
# Compare model performance
model_comparison <- data.frame(
  Model = c("Logistic Regression", "Random Forest"),
  Accuracy = c(conf_matrix$overall["Accuracy"], rf_conf_matrix$overall["Accuracy"]),
  Sensitivity = c(conf_matrix$byClass["Sensitivity"], rf_conf_matrix$byClass["Sensitivity"]),
  Specificity = c(conf_matrix$byClass["Specificity"], rf_conf_matrix$byClass["Specificity"]),
  F1_Score = c(conf_matrix$byClass["F1"], rf_conf_matrix$byClass["F1"]),
  AUC = c(auc_value, rf_auc_value)
)

# Display comparison
model_comparison %>%
  kable(caption = "Model Performance Comparison", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# 10. Feature Effects Analysis

## 10.1 Logistic Regression Effects

Let's visualize the effects of significant predictors in our logistic regression model:

```{r logistic-effects}
# Extract coefficients
coef_summary <- summary(logistic_model)$coefficients
significant_vars <- rownames(coef_summary)[coef_summary[, "Pr(>|z|)"] < 0.05]

# Format odds ratios for significant variables
sig_odds_ratios <- odds_ratios_df[odds_ratios_df$Variable %in% significant_vars, ]
sig_odds_ratios <- sig_odds_ratios[order(sig_odds_ratios$OddsRatio), ]

# Plot odds ratios for significant variables (excluding intercept)
ggplot(sig_odds_ratios[sig_odds_ratios$Variable != "(Intercept)", ], 
                   aes(x = reorder(Variable, OddsRatio), y = OddsRatio)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Odds Ratios for Significant Predictors",
       x = "Variable",
       y = "Odds Ratio (log scale)") +
  scale_y_log10() +
  theme_minimal()
```

## 10.2 Random Forest Partial Dependence Plots

Let's examine the partial dependence plots for key variables in our random forest model:

```{r partial-dependence, message=FALSE, warning=FALSE}
# Create partial dependence plot for tenure
pdp_tenure <- partial(rf_model, pred.var = "tenure_in_months", plot = TRUE, rug = TRUE, 
                     train = rf_data_train)
plot(pdp_tenure, main = "Partial Dependence on Tenure")

# Create partial dependence plot for monthly charge
pdp_charge <- partial(rf_model, pred.var = "monthly_charge", plot = TRUE, rug = TRUE, 
                     train = rf_data_train)
plot(pdp_charge, main = "Partial Dependence on Monthly Charge")

# Create partial dependence plot for service count
pdp_services <- partial(rf_model, pred.var = "service_count", plot = TRUE, rug = TRUE, 
                       train = rf_data_train)
plot(pdp_services, main = "Partial Dependence on Service Count")
```

# 11. Churn Risk Profiling

## 11.1 Predicting Churn Probability

Let's use our random forest model (which had better performance) to predict churn probability for all customers:

```{r churn-probability}
# Use the random forest model (better performance)
all_predictions_prob <- predict(rf_model, model_data, type = "prob")[, "Yes"]
model_data$churn_probability <- all_predictions_prob

# Create risk segments
model_data$risk_segment <- cut(model_data$churn_probability, 
                              breaks = c(0, 0.3, 0.6, 1),
                              labels = c("Low Risk", "Medium Risk", "High Risk"))

# Count customers in each risk segment
risk_counts <- table(model_data$risk_segment)
print(risk_counts)

# Visualize risk segments
ggplot(model_data, aes(x = risk_segment, fill = risk_segment)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  labs(title = "Customer Distribution by Churn Risk Segment",
       x = "Risk Segment",
       y = "Number of Customers") +
  scale_fill_manual(values = c("#2e6083", "#2e6083", "#52664b")) +
  theme_minimal()
```

## 11.2 High Risk Customer Profile

Let's analyze the characteristics of high-risk customers:

```{r high-risk-profile}
# Analyze high risk customer profiles
high_risk_profile <- model_data %>%
  filter(risk_segment == "High Risk") %>%
  summarize(
    count = n(),
    avg_tenure = mean(tenure_in_months),
    avg_monthly_charge = mean(monthly_charge),
    pct_month_to_month = mean(contract == "Month-to-Month") * 100,
    pct_fiber = mean(internet_type == "Fiber Optic") * 100,
    pct_no_online_security = mean(online_security == "No") * 100,
    pct_no_tech_support = mean(premium_tech_support == "No") * 100,
    avg_service_count = mean(service_count)
  )

# Display high risk profile
high_risk_profile %>%
  t() %>% 
  as.data.frame() %>%
  rownames_to_column("Metric") %>%
  setNames(c("Metric", "Value")) %>%
  kable(caption = "High Risk Customer Profile", digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# 12. Conclusion and Recommendations

## 12.1 Key Findings

Based on our comprehensive analysis, we've identified several key factors that significantly predict customer churn:

1. **Tenure**: Shorter customer tenure is strongly associated with higher churn probability.
2. **Contract Type**: Month-to-month contracts have a significantly higher churn rate (45.8%) compared to one-year (10.7%) and two-year contracts (2.5%).
3. **Internet Type**: Fiber optic internet customers have the highest churn rate.
4. **Service Adoption**: Customers without online security and tech support are more likely to churn.
5. **Monthly Charges**: Higher monthly charges are associated with increased churn risk.

## 12.2 Model Performance

Our random forest model achieved strong predictive performance:

- Accuracy: `r round(rf_conf_matrix$overall["Accuracy"], 3)`
- AUC: `r round(rf_auc_value, 3)`
- Sensitivity (True Positive Rate): `r round(rf_conf_matrix$byClass["Sensitivity"], 3)`
- Specificity (True Negative Rate): `r round(rf_conf_matrix$byClass["Specificity"], 3)`

## 12.3 Business Recommendations

Based on our findings, we recommend the following retention strategies:

1. **Target Month-to-Month Customers**: Implement targeted campaigns to convert month-to-month customers to longer-term contracts.
2. **Early Tenure Focus**: Develop specialized retention programs for customers in their first 12 months of service.
3. **Service Bundle Incentives**: Encourage adoption of online security and tech support services, which are associated with lower churn rates.
4. **Fiber Optic Customer Support**: Address potential service quality issues for fiber optic internet customers.
5. **High-Value Customer Retention**: Create specialized retention programs for customers with high monthly charges but low service adoption.

## 12.4 Implementation Plan

1. **Risk Segmentation**: Use the model to score all customers and implement tiered retention strategies.
2. **Proactive Outreach**: Contact high-risk customers before they churn with personalized offers.
3. **Service Quality Improvement**: Address potential service issues for high-churn segments.
4. **Contract Conversion Campaigns**: Offer incentives for month-to-month customers to upgrade to longer contracts.
5. **Service Bundle Promotions**: Create attractive bundles including the protective services identified in our analysis.

```{r save-high-risk}
# Save the high risk customer list for targeted interventions
# First check if customer_id exists in the original dataset
if ("customer_id" %in% colnames(telecom_churn)) {
  # Option 1: Join back the customer_id column to the high-risk customers
  high_risk_customers <- model_data %>%
    filter(risk_segment == "High Risk") %>%
    # Create a row number to join with
    mutate(row_id = row_number()) %>%
    # Add the customer_id from the original data
    left_join(
      telecom_churn %>% 
        select(customer_id) %>%
        mutate(row_id = row_number()),
      by = "row_id"
    ) %>%
    # Remove the temporary row_id
    select(-row_id) %>%
    # Now select the desired columns
    select(customer_id, churn_probability, tenure_in_months, monthly_charge, 
           contract, internet_type, online_security, premium_tech_support)
} else {
  # Option 2: If there's no customer_id at all, create a sequential ID
  high_risk_customers <- model_data %>%
    filter(risk_segment == "High Risk") %>%
    # Create a sequential ID
    mutate(customer_id = paste0("HR_", row_number())) %>%
    select(customer_id, churn_probability, tenure_in_months, monthly_charge, 
           contract, internet_type, online_security, premium_tech_support)
}

# Check the result
head(high_risk_customers)

# Write the high risk customers to a CSV file
write.csv(high_risk_customers, "high_risk_customers.csv", row.names = FALSE)
```

---